\bf Estimating the number of clusters using Cross Validation

### 2016-04-20 New simulation/technometrics.tex
Details about enrichment anlaysis to aid understanding of table 2

\cite{tavazoie1999systematic} applied $k$-means clustering on the preprocessed data with $k=30$ and discovered several periodic clusters. The preprocessing procedure includes choosing the most variable $2,945$ ORFs among the total $6,220$ by metric of variation based on the normalized dispersion across time points (s.d/mean). Data at time points $90$ and $100$ minutes were removed because of their less efficient labelling of mRNA during the original chip hybridization. It results a $2,945$ by $15$ data matrix. The last step of the preprocessing procedure is to variance normalize each gene across the $15$ time points, which is done by subtracting the mean and then dividing by the standard deviation. The result is each gene has mean zero and standard deviation equal to one across the time points.

@@

We also applied our proposed Gabriel method on the preprocessed data of \cite{tavazoie1999systematic}. The algorithm found $5$ clusters. Since the dimensions (time points) are likely to be correlated, we also applied the correlation-corrected version of Gabriel method on the data which also result $5$ clusters. This numbers is much smaller compare to those of \cite{tavazoie1999systematic} and \cite{dortet2008model}. We will focus our analysis on this $5$ clusters found by the proposed method, whose result is summarized in Figure \ref{fig:5_clusters}, where each cluster is presented by the mean expression level of all the genes within that cluster over $15$ time points. Similar as \cite{tavazoie1999systematic}, we found the members of each cluster to be significantly enriched for genes function in similar biological process. Each gene is mapped with the Gene Ontology Annotation in SGD database, which categorizes the genes into $103$ groups in terms of biological process. The $p$-value of observing $k$ genes from a particular category group within a cluster of size $m$ is given by:
\[	P = 1-\sum^{k-1}_{i=1} \frac{{{2945-n}\choose{m-i}}{{n}\choose{i}} }{ {{2945} \choose{m}} }\] 
where $n$ is the total ORFs in that group. The enrichment of each category is shown in Table \ref{table:enrich}.

\begin{table}
\begin{center}
\captionsetup{justification=centering}
\caption{\label{table:enrich} Enrichment of clusters for ORFs with biological process categories}
\scalebox{0.8}{%
\begin{tabular}{cclcl}
\hline
   Cluster & Cluster size($m$) & Biological process category(total ORFs) & ORFs within category($k$) &$p$-value \\ \hline
 $1$ & $550$  &  response to oxidative stress ($55$)  &    $24$  & $1.5\times10^{-5}$ \\
     &     &  response to chemical ($213$)    &    $64$     & $2.2\times10^{-5}$\\
  $2$& $590$  &  mitochondrion organization ($159$) & $79$ &  $1.1\times10^{-16}$  \\  
  	 &     &  mitochondrial translation ($51$) & $28$ 	&  $2.9\times10^{-8}$  \\  
  	 &     &  generation of precursor metabolites and energy ($80$) &  $37$ & $7.3\times10^{-8}$   \\  
 $3$ & $654$ &  transcription from RNA polymerase II promoter ($214$) & $75$ &  $5.5\times10^{-6}$    \\  
  	 &     & mRNA processing ($67$) & $30$ 	& $2.7\times10^{-5}$    \\  
  	 &     & mitotic cell cycle ($183$)  & $63$ 	& $6.2\times10^{-5}$  \\ 
  $4$& $634$  & cytoplasmic translation ($134$)  &  $105$ &  $3.3\times10^{-47}$   \\  
     &     &  ribosomal subunit biogenesis $(138)$ & $73$ 	&  $7.7\times10^{-17}$  \\  
  	 &     &  rRNA processing ($131$) &  $61$	&  $5.7\times10^{-11}$  \\  
  	 &     &  ribosome assembly ($36$) &  $21$	&  $1.5\times10^{-6}$  \\  
$5$  &   $517$   &  chromosome segregation($106$) &  $53$      &  $6.0\times10^{-15}$ \\
     &     &  cellular response to DNA damage stimulus ($172$)  &  $71$  & $3.6\times10^{-14}$     \\  
  	 &     &  DNA repair ($147$)  &  $64$  & $3.7\times10^{-14}$     \\                 
 	 &     &  DNA replication ($78$)  & $42$   &  $1.8\times10^{-13}$   \\  
  	 &     &  mitotic cell cycle ($183$)  &  $70$  &  $4.8\times10^{-12}$   \\

>>>>

\cite{tavazoie1999systematic} applied $k$-means clustering on the preprocessed data with $k=30$ and discovered several periodic clusters. The preprocessing procedure includes choosing the most variable $2,945$ geness among the total $6,220$ by metric of variation based on the normalized dispersion across time points (s.d/mean). Data at time points $90$ and $100$ minutes were removed because of their less efficient labelling of mRNA during the original chip hybridization. It results a $2,945$ by $15$ data matrix. The last step of the preprocessing procedure is to variance normalize each gene across the $15$ time points, which is done by subtracting the mean and then dividing by the standard deviation. The result is each gene has mean zero and standard deviation equal to one across the time points.

@@

We also applied our proposed Gabriel method on the preprocessed data of \cite{tavazoie1999systematic}. The algorithm found $5$ clusters. Since the dimensions (time points) are likely to be correlated, we also applied the correlation-corrected version of Gabriel method on the data which also result $5$ clusters. This numbers is much smaller compare to those of \cite{tavazoie1999systematic} and \cite{dortet2008model}. We will focus our analysis on this $5$ clusters found by the proposed method, whose result is summarized in Figure \ref{fig:5_clusters}, where each cluster is presented by the mean expression level of all the genes within that cluster over $15$ time points. Similar as \cite{tavazoie1999systematic}, we found the members of each cluster to be significantly enriched for genes function in similar biological process. Each gene is mapped with the Gene Ontology (GO) Annotation in Saccharomyces Genome Database (SGD) database, which provides Gene Ontology (GO) annotation of the entire genome using GO slims, which are high-level subsets of GO terms that allow grouping of genes into broad categories, such as biological process, molecular function or cellular component. We choose to group genes in terms of biological process, which is a recognized series of events or molecular functions.

To identify biologically meaningful gene clusters from spurious ones, \cite{tavazoie1999systematic} introduced a method, enrichment analysis, to assess the statistical significance of the over-represented gene class in a cluster using hypergeometric distribution. Specifically, the $p$-value of observing $k$ genes from a particular category group within a cluster of size $m$ is given by:
\[	P = 1-\sum^{k-1}_{i=1} \frac{{{2945-n}\choose{m-i}}{{n}\choose{i}} }{ {{2945} \choose{m}} }\] 
where $n$ is the total genes in that group. It can be seen as testing the null hypothesis that genes are randomly  distributed among each cluster. The enrichment of each category is shown in Table \ref{table:enrich}.

\begin{table}
\begin{center}
\captionsetup{justification=centering}
\caption{\label{table:enrich} Enrichment of clusters for genes within biological process categories}
\scalebox{0.8}{%
\begin{tabular}{cclcl}
\hline
   Cluster & Cluster size($m$) & Biological process category(total genes) & genes within category($k$) &$p$-value \\ \hline
 $1$ & $550$  &  response to oxidative stress ($55$)  &    $24$  & $1.5\times10^{-5}$ \\
     &     &  response to chemical ($213$)    &    $64$     & $2.2\times10^{-5}$\\
  $2$& $590$  &  mitochondrion organization ($159$) & $79$ &  $1.1\times10^{-16}$  \\  
  	 &     &  mitochondrial translation ($51$) & $28$ 	&  $2.9\times10^{-8}$  \\  
  	 &     &  generation of precursor metabolites and energy ($80$) &  $37$ & $7.3\times10^{-8}$   \\  
 $3$ & $654$ &  transcription from RNA polymerase II promoter ($214$) & $75$ &  $5.5\times10^{-6}$    \\  
  	 &     & mRNA processing ($67$) & $30$ 	& $2.7\times10^{-5}$    \\  
  	 &     & mitotic cell cycle ($183$)  & $63$ 	& $6.2\times10^{-5}$  \\ 
  $4$& $634$  & cytoplasmic translation ($134$)  &  $105$ &  $3.3\times10^{-47}$   \\  
     &     &  ribosomal subunit biogenesis $(138)$ & $73$ 	&  $7.7\times10^{-17}$  \\  
  	 &     &  rRNA processing ($131$) &  $61$	&  $5.7\times10^{-11}$  \\  
  	 &     &  ribosome assembly ($36$) &  $21$	&  $1.5\times10^{-6}$  \\  
$5$  &   $517$   &  chromosome segregation($106$) &  $53$      &  $6.0\times10^{-15}$ \\
     &     &  cellular response to DNA damage stimulus ($172$)  &  $71$  & $3.6\times10^{-14}$     \\  
  	 &     &  DNA repair ($147$)  &  $64$  & $3.7\times10^{-14}$     \\                 
 	 &     &  DNA replication ($78$)  & $42$   &  $1.8\times10^{-13}$   \\  
  	 &     &  mitotic cell cycle ($183$)  &  $70$  &  $4.8\times10^{-12}$   \\
